Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-04-14 WSSSPE2_report/WSSSPE2_review.tex
adding some travel support stuff in intro, finishing 8.5 (prizes)

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
the development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational, and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged the broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future practice for software development and usage in science and
engineering. The workshop included multiple mechanisms for participation,
encouraged team building around solutions, and identified risky solutions with
potentially transformative outcomes. It strongly encouraged participation of
early-career scientists, postdoctoral researchers, and graduate students.
\todo{record how/number of  travel awards were created \& recipients selected?}

@@

Some papers explored gaps in training of early career scientists. Industry
participants in our group confirm this. We asked ourselves, ``Are
traditional courses failing?''. We think yes. Changes to undergraduate
curricula requirements are difficult. But as programming models become
more complex, we have to raise the skill levels. Skills must be improved
not only in traditional programming - learning languages and algorithms -
but around professional software development. We want to get to a point
where ``Everyone has a new minimum now - everyone knows git''. Developers
also need training on licensing choices. Just because a code is on git
doesn't mean it's open.

@@

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?''. One example demonstrates
change over time. In 2007, nanoHUB listed the academic reward structure as
a problem in an EDUCAUSE report where the authors note, ``In the future,
nanoHUB researchers are hoping to change the research culture. While they
recognize that young faculty members are unlikely to get tenure based on
their nanoHUB contributions, they hope to encourage faculty to think
beyond their own research needs to consider publishing tutorials and other
Web content in their field on the nanoHUB site''. Fast forward seven years
later where, in a 2014 iSGTW article, quote nanoHUB PI Gerhard Klimeck, ``A
former student of mine published eight tools on nanoHUB, serving over
6,000 people with his tools. He then joined a university as a professor
and introduced nanoHUB. Use of the gateway from that university
skyrocketed; he used nanoHUB in existing classes, created new classes, and
infused it in his research. Ultimately, the professor's department head
associates his two-year rise to tenure with the notoriety and innovation
he gained through nanoHUB.''

The group also discussed ``attribution trees'', an idea put forward by Dan
Katz and Arfon Smith where a chain of attributions can, for example, give
credit to those developing libraries and building blocks that other pieces
of software use. The group considered where to push the use of this, which
journals? On participant noted that Dryad works with the journals. If a
paper is accepted, the supporting data must be submitted. We discussed
reproducibility as a component of the journal review process and ``active
papers'', with immediate links to the data and software.

@@

We then moved beyond our training discussion to address funding that
encourages sustainable software -- funding of both people and of projects
that create a true system to support sustainability. We called this
``institutionalized serendipity''. Because science is increasingly reliant
on software, one participant observed that ``software development can have
much broader impact than publishing individual research, but it is not
viewed that way''. Because of this centrality, one participant mentioned
that training ought to be called Science Carpentry rather than Software
Carpentry. We felt we were beginning to see changes in the research
community as a result of the NSF's data management plan requirements. PIs
are thinking more about data and some university libraries are offering
data repositories. We wondered if a software management plan might be
effective. "How might funding programs need to adapt to reward good
software development practices?'', we asked ourselves. We thought about
best practices such as version control, test harnesses, mailing lists, bug
tracking, community contribution, and reuse where appropriate. We thought
about measuring success through usage statistics (downloads, altmetrics).
"Should funders demand reproducibility?'' we wondered? In order for results
to be reproduced, software would need to be carefully curated.

Again, our industry participants contributed unique viewpoints. Partnering
with industry was seen one path to sustainability. The unique partnerships
Kitware engages in promote academic freedom while creating an income
stream from certain portions of the software. This type of approach to
sustainable software frees researchers from performing tasks that do not
offer the rewards their institution values''. We talked some about
successful models for industry partnerships that preserve open science.
Participants noted that there are some NSF programs that prohibit
partnerships with for profit companies (but there are others where this is
encouraged).

@@

\katznote{I don't understand this at all.  Can someone explain it?}
There are also other indirect forms of incentives
(though these are not a direct prize, they still can incentivize someone):
\begin{itemize}
\item Converting reputation by joining the Apache Software Foundation, Google, etc. 
\item Inviting ASF and open source people to contribute to scientific code (at
least on questions of overlapping
\item Template for assessing scientific contributions made through software.   
\end{itemize}

To determine who should give prizes, perhaps we should find those who we think would
be awarded prizes, then ask them who they would want to receive a prize from, then
ask those organizations to see if they are willing to be involved in the process.

One idea is to create a funding program for disciplines or other organizations to
create a prize program.  We provide a framework, for example: awards to individuals, must
award in 5 or 6 areas, must have a jury and/or objective criteria that includes
senior/junior domain people and technology people, must have the recipients awarded
at a relevant event, must provide citations that explain why the prizes should be awarded.
A concern is that having many organizations award prizes may reduce the impact.

Possible awarders are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology
Society of America, etc. Or perhaps this should be a joint technology/science
partnership, for example, the [Apache $|$ Mozilla]--[AAAS $|$ disciplinary
society] prize?

Some potential criteria for prizes are:
community engagement, helping out others;
number of unique contributors;
adding new pieces of functionality to software;
integrating software into broader ecosystem; championing broad principles of
sustainability, open science, open source, etc.;
improving accessibility to software, to scientific software (perhaps championing
inclusiveness or making software accessible?);
documentation;
tutorials;
commits / patching;
leadership transition; and
best contribution by a non-founder.

Perhaps there should be different criteria for ``established'' members of the
community versus junior members? Or perhaps prizes should be restricted to
junior members?

An important point is that we don't want to give prizes just to reward people
who are really good at this one thing, but rather we want to reward people who
are building the culture we want as scientists.

@@

\choinote{Add missing speaker names} \katznote{they are all there}

>>>>

Progress in scientific research is dependent on the quality and accessibility of
software at all levels. Hence it is critical to address challenges related to
the development, deployment, maintenance, and overall sustainability of reusable
software as well as education around software practices. These challenges can be
technological, policy based, organizational, and educational, and are of
interest to developers (the software community), users (science disciplines),
software-engineering researchers, and researchers studying the conduct of
science (science of team science, science of organizations, science of science
and innovation policy, and social science communities). The WSSSPE1 workshop
engaged the broad scientific community to identify challenges and best practices
in areas of interest to creating sustainable scientific software. WSSSPE2
invited the community to propose and discuss specific mechanisms to move towards
an imagined future practice for software development and usage in science and
engineering. The workshop included multiple mechanisms for participation,
encouraged team building around solutions, and identified risky solutions with
potentially transformative outcomes. It strongly encouraged participation of
early-career scientists, postdoctoral researchers, and graduate students, with
funds provided to the conference organizers by the Moore Foundation and the
National Science Foundation, to support the travel of potential participants who
would not otherwise be able to attend, and young participants and those from
underrepresented groups, respectively.  These funds allowed \todo{number}
additional participants to attend, and each was offered the chance to present a
lightning talk.

@@

Some papers explored gaps in training of early career scientists. Industry
participants in our group confirm this. We asked ourselves, ``Are
traditional courses failing?'' We think yes. Changes to undergraduate
curricula requirements are difficult. But as programming models become
more complex, we have to raise the skill levels. Skills must be improved
not only in traditional programming - learning languages and algorithms -
but around professional software development. We want to get to a point
where ``Everyone has a new minimum now - everyone knows git.'' Developers
also need training on licensing choices. Just because a code is on git
doesn't mean it's open.

@@

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?'' One example demonstrates
change over time. In 2007, nanoHUB listed the academic reward structure as
a problem in an EDUCAUSE report where the authors note, ``In the future,
nanoHUB researchers are hoping to change the research culture. While they
recognize that young faculty members are unlikely to get tenure based on
their nanoHUB contributions, they hope to encourage faculty to think
beyond their own research needs to consider publishing tutorials and other
Web content in their field on the nanoHUB site.'' Fast forward seven years
later where, in a 2014 iSGTW article, quote nanoHUB PI Gerhard Klimeck, ``A
former student of mine published eight tools on nanoHUB, serving over
6,000 people with his tools. He then joined a university as a professor
and introduced nanoHUB. Use of the gateway from that university
skyrocketed; he used nanoHUB in existing classes, created new classes, and
infused it in his research. Ultimately, the professor's department head
associates his two-year rise to tenure with the notoriety and innovation
he gained through nanoHUB.''

The group also discussed ``attribution trees,'' an idea put forward by Dan
Katz and Arfon Smith where a chain of attributions can, for example, give
credit to those developing libraries and building blocks that other pieces
of software use. The group considered where to push the use of this, which
journals? On participant noted that Dryad works with the journals. If a
paper is accepted, the supporting data must be submitted. We discussed
reproducibility as a component of the journal review process and ``active
papers,'' with immediate links to the data and software.

@@

We then moved beyond our training discussion to address funding that
encourages sustainable software -- funding of both people and of projects
that create a true system to support sustainability. We called this
``institutionalized serendipity.'' Because science is increasingly reliant
on software, one participant observed that ``software development can have
much broader impact than publishing individual research, but it is not
viewed that way.'' Because of this centrality, one participant mentioned
that training ought to be called Science Carpentry rather than Software
Carpentry. We felt we were beginning to see changes in the research
community as a result of the NSF's data management plan requirements. PIs
are thinking more about data and some university libraries are offering
data repositories. We wondered if a software management plan might be
effective. "How might funding programs need to adapt to reward good
software development practices?'' we asked ourselves. We thought about
best practices such as version control, test harnesses, mailing lists, bug
tracking, community contribution, and reuse where appropriate. We thought
about measuring success through usage statistics (downloads, altmetrics).
"Should funders demand reproducibility?'' we wondered. In order for results
to be reproduced, software would need to be carefully curated.

Again, our industry participants contributed unique viewpoints. Partnering
with industry was seen one path to sustainability. The unique partnerships
Kitware engages in promote academic freedom while creating an income
stream from certain portions of the software. This type of approach to
sustainable software frees researchers from performing tasks that do not
offer the rewards their institution values.'' We talked some about
successful models for industry partnerships that preserve open science.
Participants noted that there are some NSF programs that prohibit
partnerships with for profit companies (but there are others where this is
encouraged).

@@

Other (non-prize) forms of incentives are
\begin{itemize}
\item Converting reputation by joining the Apache Software Foundation, Google, etc. 
\item Inviting ASF and open source people to contribute to scientific code (at
least on questions of overlapping
\item Template for assessing scientific contributions made through software.   
\end{itemize}

The group suggested that to determine who should give out prizes, perhaps we should find those who we think would
be awarded prizes, then ask them who they would want to receive a prize from, then
ask those organizations to see if they are willing to be involved in the process.

One of the group's ideas was to create a funding program for disciplines or other organizations to
create a prize program.  We would provide a framework and a set of requirements,
for example: awards to individuals; must
award in 5 or 6 areas; must have a jury and/or objective criteria that includes
senior/junior domain people and technology people; must have the recipients awarded
at a relevant event; and must provide citations that explain why the prizes should be awarded.
Different organizations could then decide to sign up to the framework and give out
awards under this general brand.  However, there was a concern that having many
organizations award prizes may reduce the impact of the prizes.

Possible groups that might give out awards, either under our framework, or more generally,
are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology
Society of America, etc. Perhaps this could be a joint technology/science
partnership, for example, the [Apache $|$ Mozilla]--[AAAS $|$ disciplinary
society] prize.

Some potential criteria for prizes the group suggested are:
community engagement, helping out others;
number of unique contributors;
adding new pieces of functionality to software;
integrating software into broader ecosystem; championing broad principles of
sustainability, open science, open source, etc.;
improving accessibility to software, to scientific software (perhaps championing
inclusiveness or making software accessible?);
documentation;
tutorials;
commits / patching;
leadership transition; and
best contribution by a non-founder.

The group discussion if there should be different criteria for ``established'' members of the
community versus junior members, and if prizes should be restricted to
junior members, but left these as open questions.

A point the group considered important is that we don't want to give prizes just to reward people
who are really good at this one thing, but rather we want to reward people who
are building the culture we want as scientists.

@@
